\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-14 ace-pldi14/ace-pldi14.tex
pldi14 2.5 p2+3 updated

\subsection{Generic Functions}
\begin{codelisting}
\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
\label{map}
\end{codelisting}
\begin{codelisting}[t]
\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
\label{metaprogramming}
\end{codelisting}
Lines 3-4 introduce \verb|map|, an Ace function governed by the active base \verb|clx.base| taking three arguments. The \emph{active target}, \verb|clx.opencl|, determines which language the function will target (here, the OpenCL kernel language) and mediates code generation.

The body of this function, highlighted in grey for emphasis, does not have Python's semantics. Instead, it will be governed by the active base together with the active types used within it. No such types have been provided explicitly for the arguments yet, however. Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments, so \verb|map| is not yet a \emph{concrete function}. We call functions awaiting types \emph{generic functions}. %Once types have been assigned, they are called \emph{concrete functions}.

@@

\subsection{Concrete Functions and Explicit Compilation}
To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, so that typechecking and translation can proceed. Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| via the subscript operator. We attempt to do so three times in Listing \ref{compscript}. The first, on line \ref{compscript}.7, fails due to a type error, which we handle so that the script can proceed. The error occurred  because the ordering of types was incorrect. We provide a valid ordering on line \ref{compscript}.9 to generate the concrete function \verb|map_neg_f32|. We then provide a different type assignment to generate the concrete function \verb|map_neg_ci32|.
Concrete functions are instances of \verb|ace.ConcreteFn|, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function. %The typing and translation process is mediated by the logic in the active base, types and target that have been provided, as we will describe in more detail below.

To produce an output file from an Ace compilation script like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}. The \verb|acec| compiler operates in two steps:
\begin{enumerate}
\item Executes the provided Python file (\verb|listing3.py|).
\item Extracts the translation from concrete functions and other top-level constructs (e.g. types requiring declarations, or generated imports and pragmas) in the top-level Python environment.  This may produce one or more files (here, just \verb|listing3.cl|) as determined by the active targets that were used (for example, a web framework built upon Ace might produce separate HTML, CSS and JavaScript files; see Sec. \ref{atargets}).
\end{enumerate}

In this case, step 1 prints the output on lines \ref{mapc}.2-\ref{mapc}.4. The type error printed on lines \ref{mapc}.3-\ref{mapc}.4 will be explained in the next section. The compiler (which is a simple Python script) concludes with the message on line \ref{mapc}.5 to indicate that one file was generated. This file is shown in Listing \ref{mapout} and can be used by any programs that consume OpenCL code (e.g. a C file that invokes the generated kernels via the OpenCL host API). 
We will show in Section \ref{backend} that for targets with Python bindings, such as OpenCL, CUDA, C, Java or Python itself, generic functions can be executed directly, without any of the explicit compilation steps in Listings \ref{compscript}-\ref{mapc}.

\subsection{Types}
\begin{codelisting}
\lstinputlisting{listing4.py}
\caption{[\texttt{listing\ref{compscript}.py}] The generic \texttt{map} function compiled to map the \texttt{negate} function over two  types of input.}
\label{compscript}
\end{codelisting}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing3.py`
Hello, compile-time world!
[ace] TypeError in listing1.py (line 6, col 28): 
      'GenericFnType(negate)' does not support [].
[acec] listing3.cl successfully generated.
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
\label{mapc}
\end{codelisting}
\begin{codelisting}
\lstinputlisting[style=OpenCL]{listing5.cl}
\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
\label{mapout}
\end{codelisting}
Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned on lines \ref{compscript}.7-\ref{compscript}.10. In Ace, types are themselves values that can be manipulated at compile-time, in contrast to most  contemporary languages, where user-defined types are declaratively written at compile-time but cannot be constructed, inspected or passed around programmatically. More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec. \ref{atypes}). Python, then, serves as the \emph{type-level language} of Ace. 
As Python values, types can be assigned to variables when desired (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). Types, like all compile-time objects derived from Ace base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.

The type named \verb|T1| on line \ref{compscript}.3 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL \cite{opencl11}). It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to a value representing an  address space, \verb|clx.global_|, and the type being pointed to. That type, \verb|clx.float|, is in turn the Ace type corresponding to \verb|float| in OpenCL (which, unlike C99, is always 32 bits). 
The \verb|clx| library contains a full implementation of the OpenCL type system.
We note that we are being more verbose than necessary for the sake of pedagogy. The \verb|clx| library includes a concise shorthand for OpenCL's built-in types: \verb|T1| could be written \verb|clx.gp(clx.f32)|. We also note that Ace is \emph{unopinionated} about issues like memory safety. We will show how to implement safer alternatives to raw pointers in Sec. \ref{examples}\todo{true?}, but we allow users to choose a low level of abstraction if the need arises.

The type \verb|T2| on line \ref{compscript}.4 does not correspond to type in OpenCL directly. It is a pointer to a \emph{complex integer} in global memory. OpenCL does not include primitive support for complex numbers, however -- it is an extension available only in \verb|clx|. The \verb|clx.Cplx| type constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \verb|clx.int|, the type corresponding to 32-bit OpenCL integers. Arithmetic is supported with other types in the same family, as well as with plain numeric types (treated as if their imaginary part was zero). When targeting OpenCL, Ace expressions of type \verb|clx.Cplx(clx.int)| are compiled to OpenCL expressions of type \verb|int2|, a pair of 32-bit integers. This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21. This is merely an implementation detail of \verb|clx.Cplx|, however. An Ace value of type \verb|clx.int2| can\emph{not} be used when a \verb|clx.Cplx(clx.int)| is expected (attempting to do so will result in a static type error). That is, \verb|clx.Cplx| truly extends the type system, it is not merely a type alias (see Sec. \ref{atypes}).

The type \verb|TF| on line \ref{compscript}.5

>>>>

\subsection{Generic Functions}\label{genfn}
\begin{codelisting}
\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
\label{map}
\end{codelisting}
\begin{codelisting}[t]
\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
\label{metaprogramming}
\end{codelisting}
Lines 3-4 introduce \verb|map|, an Ace function of three arguments that is governed by the \emph{active base} referred to by \verb|clx.base| and targeting the \emph{active target} referred to by \verb|clx.opencl|. The active target determines which language the function will compile to (here, the OpenCL kernel language) and mediates code generation.

The body of this function, highlighted in grey for emphasis, does not have Python's semantics. Instead, it will be governed by the active base together with the active types used within it. No such types have been provided explicitly, however. Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments. We call functions awaiting types \emph{generic functions}. Once types have been assigned, they are called \emph{concrete functions}.

@@

\subsection{Concrete Functions and Explicit Compilation}
To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, and typechecking and translation must then succeed. Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| using the subscript operator. We attempt to do so three times in Listing \ref{compscript}. The first, on line \ref{compscript}.7, fails due to a type error, which we handle so that the script can proceed. The error occurred  because the ordering of the argument types was incorrect. We provide a valid ordering on line \ref{compscript}.9 to generate the concrete function \verb|map_neg_f32|. We then provide a different type assignment to generate the concrete function \verb|map_neg_ci32|.
Concrete functions are instances of \verb|ace.ConcreteFn|, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function. %The typing and translation process is mediated by the logic in the active base, types and target that have been provided, as we will describe in more detail below.

To produce an output file from an Ace ``compilation script'' like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}. The \verb|acec| compiler (a simple Python script) operates in two stages:
\begin{enumerate}
\item Executes the provided Python file (\verb|listing3.py|).
\item Extracts the translations from concrete functions and other top-level constructs (e.g. types requiring declarations, or generated imports and pragmas) in the top-level Python environment.  This may produce one or more files as mediated by the active targets that were used (here, just \verb|listing3.cl|, but a web framework built upon Ace might produce separate HTML, CSS and JavaScript files; see Sec. \ref{atargets}).
\end{enumerate}

In this case, stage 1 results in the output on lines \ref{mapc}.2-\ref{mapc}.4. The type error printed on lines \ref{mapc}.3-\ref{mapc}.4 will be explained in the next section. The compiler then enters stage 2 and concludes with the message on line \ref{mapc}.5 to indicate that one file was generated. This file is shown in Listing \ref{mapout} and can be used by any programs that consume OpenCL code (e.g. a C program that invokes the generated kernels via the OpenCL host API). 
We will show in Section \ref{backend} that for targets with Python bindings, such as OpenCL, CUDA, C, Java or Python itself, generic functions can be executed directly, without any of the explicit compilation steps in Listings \ref{compscript}-\ref{mapc}.

\subsection{Types}
\begin{codelisting}
\lstinputlisting{listing4.py}
\caption{[\texttt{listing\ref{compscript}.py}] The generic \texttt{map} function compiled to map the \texttt{negate} function over two  types of input.}
\label{compscript}
\end{codelisting}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing3.py`
Hello, compile-time world!
[ace] TypeError in listing1.py (line 6, col 28): 
      'GenericFnType(negate)' does not support [].
[acec] listing3.cl successfully generated.
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
\label{mapc}
\end{codelisting}
\begin{codelisting}
\lstinputlisting[style=OpenCL]{listing5.cl}
\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
\label{mapout}
\end{codelisting}
Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned to the arguments of \verb|map| on lines \ref{compscript}.7-\ref{compscript}.10. In Ace, types are themselves values that can be manipulated at compile-time. This stands in contrast to other contemporary languages, where user-defined types (e.g. datatypes, classes, structs) are written declaratively at compile-time but cannot be constructed, inspected or passed around programmatically. More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec. \ref{atypes}). 
As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). Types, like all compile-time objects derived from Ace base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.

The type named \verb|T1| on line \ref{compscript}.3 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four address spaces defined by OpenCL \cite{opencl11}). It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to a value representing the  address space, \verb|clx.global_|, and the type being pointed to. That type, \verb|clx.float|, is in turn the Ace type corresponding to \verb|float| in OpenCL (which, unlike C99, is always 32 bits). 
The \verb|clx| library contains a full implementation of the OpenCL type system (including behaviors, like promotions, inherited from C99).
Ace is \emph{unopinionated} about issues like memory safety and the wisdom of such promotions. We will discuss how to implement, as libraries, abstractions that are higher-level than raw pointers in Sec. \ref{examples}, but Ace does not prevent users from choosing a low level of abstraction or ``interesting'' semantics if the need arises (e.g. for compatibility with existing libraries; see the discussion in Sec. \ref{discussion}). We also note that we are being more verbose than necessary for the sake of pedagogy. The \verb|clx| library includes more concise shorthand for OpenCL's types: \verb|T1| is equal to \verb|clx.gp(clx.f32)|. %Similarly, the decorators in Listings \ref{map} and \ref{metaprogramming} could have been written \verb|clx.cl_fn|.\todo{move this back there probably}

The type \verb|T2| on line \ref{compscript}.4 is a pointer to a \emph{complex integer} in global memory. It does not correspond direrctly to a type in OpenCL, because OpenCL does not include primitive support for complex numbers. Instead, it uses an active type constructor \verb|clx.Cplx|, which includes the necessary logic for typechecking operations on complex numbers and translating them to OpenCL (Sec. \ref{atypes}). This constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \verb|clx.int|, which corresponds to 32-bit OpenCL integers. Arithmetic operations with other complex numbers, as well as with plain numeric types (treated as if their imaginary part was zero), are supported. When targeting OpenCL, Ace expressions assigned type \verb|clx.Cplx(clx.int)| are compiled to OpenCL expressions of type \verb|int2|, a  \emph{vector type} of two 32-bit integers (a type that itself is not inherited from C99). This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21. This choice is merely an implementation detail that can be kept private to \verb|clx|, however. An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error).% \verb|clx.Cplx| truly extends the type system, it is not a type alias.

The type \verb|TF| on line \ref{compscript}.5 is extracted from the generic function \verb|negate| constructed in Listing \ref{metaprogramming}. Generic functions, according to Sec. \ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to it.
